### 2020-09-04 paper/coloringBergeGraphs.tex
Add citation
Work on combinatorial graphs

\begin{wrapfigure}{r}{0.35\textwidth}
  \label{fig:c5}
	\input{tikzpictures/c5.tex}
	\caption{Example noisy channel}%
	% \vspace{-0.5cm}
\end{wrapfigure}

@@

For a channel defined by graph $C_5$ using five messages of length 2 gives us a lower bound on $\Theta(G)$ equal $\sqrt{5}$ (as does calculating $\alpha(C_5^2)$).

Consider an "umbrella" in $\mathbb{R}^3$ with the unit vector $e_1 = (1, 0, 0)$ as its "handle" and 5 "ribs" of unit length \todo{picture}. Open it up to the point where non-consecutive ribs are orthogonal, that is form an angle of 90°. This way we get a representation of $C_5$ by 5 unit vectors $u_1, \ldots u_5$ so that each $u_i$ forms the same angle with $e_1$ and any two non-adjacent nodes are represented with orthogonal vectors. We can calculate $e_1^\intercal u_i = 5 ^ {-1/4}$.

It turns out, that we can obtain a similar representation of the nodes of $C_5^k$ by unit vectors $v_i \in \mathbb{R}^3k$, so that any two non-adjacent nodes are labeled with orthogonal vectors (this representation is sometimes called the \emph{orthogonal representation} \cite{Lovsz1989Orthogonal}). Moreover, we still get $e_1^\intercal v_i = 5^{-k/4}$ for every $i \in V(C_5^k)$ (the proof is quite technical and we omit it here).

@@

Contrary to \Lovasz's first hope \cite{Lovasz1979} $\vartheta(G)$ does not always equal $\Theta(G)$, it is only an upper bound on Shannon capacity of the graph. However, these two equal in some graphs, including all perfect graphs, as is demonstrated in the \Lovasz "sandwich theorem" \cref{thm:sandwich}.

@@

\subsection{Coloring perfect graph using ellipsoid method}
\subsubsection{Maximum cardinality stable set}

\TODO{cite the paper on using theta to color.}

@@

>>>>

\begin{wrapfigure}{r}{0.35\textwidth}
	\input{tikzpictures/c5.tex}
	\caption{Example noisy channel}%
  \label{fig:c5}
	% \vspace{-0.5cm}
\end{wrapfigure}

@@

For a channel defined by a graph $C_5$, using five messages of length 2 to communicate gives us a lower bound on $\Theta(G)$ equal $\sqrt{5}$ (as does calculating $\alpha(C_5^2)$).

Consider an "umbrella" in $\mathbb{R}^3$ with the unit vector $e_1 = (1, 0, 0)$ as its "handle" and 5 "ribs" of unit length \todo{picture}. Open it up to the point where non-consecutive ribs are orthogonal, that is form an angle of 90$^\circ$. This way we get a representation of $C_5$ by 5 unit vectors $u_1, \ldots u_5$ so that each $u_i$ forms the same angle with $e_1$ and any two non-adjacent nodes are represented with orthogonal vectors. We can calculate $e_1^\intercal u_i = 5 ^ {-1/4}$.

It turns out, that we can obtain a similar representation of the nodes of $C_5^k$ by unit vectors $v_i \in \mathbb{R}^{3k}$, so that any two non-adjacent nodes are labeled with orthogonal vectors (this representation is sometimes called the \emph{orthogonal representation} \cite{Lovsz1989Orthogonal}). Moreover, we still get $e_1^\intercal v_i = 5^{-k/4}$ for every $i \in V(C_5^k)$ (the proof is quite technical and we omit it here).

@@

Contrary to \Lovasz's first hope \cite{Lovasz1979} $\vartheta(G)$ does not always equal $\Theta(G)$, it is only an upper bound on it. However, these two equal in some graphs, including all perfect graphs, as is demonstrated in the \Lovasz "sandwich theorem" (\cref{thm:sandwich}).

@@

\subsection{Coloring perfect graph using ellipsoid method}
The following is based on \citetitle{Laurent2005} by \citeauthor{Laurent2005} \cite{Laurent2005}.

\subsubsection{Maximum cardinality stable set}

@@

Ever since Grötschel et al. published an ellipsoid-method-based polynomial algorithm for coloring perfect graphs, a combinatorial algorithm for doing the same has been sought. As of yet, it is not known, although there is much progress in the field.

A \emph{prism} is a graph consisting of two disjoint triangles and two disjoint paths between them. Notice, that for a graph to contain no odd hole, all three paths in a prism must have the same parity. A prism with all three paths odd is called an \emph{odd prism}.

In 2005 Maffray and Trotignon a coloring algorithm that colors graphs containing no odd hole, no antihole and no prism (sometimes called Artemis graphs) in $O(|V|^4|E|)$ time \cite{Maffray2006}. They later improved the time complexity to $O(|V|^2|E|)$ \cite{Lvque2009}.

In 2015 Maffray showed an algorithm for coloring Berge graphs with no squares (a square is a $C_4$) and no odd prism \cite{Maff2015}.

In 2016 Chudnovsky et al. published an algorithm that given a perfect graph $G$ with $\omega(G) = k$ colors it optimally in a time polynomial for a fixed $k$ \cite{Chudnovsky2017}.

A most recent advancement (2018) is an algorithm by Chudnovsky et al. that colors any square-free Berge graphs in time of $O(|V|^9)$ \cite{Chudnovsky2019}. Before proving strong perfect graph conjecture, a similar conjecture for square-free Berge graphs has been proven by Conforti et al. \cite{Conforti2004} During one of her lectures, Maria Chudnovsky expressed hope that discovery of full algorithm for coloring Berge graphs might follow a similar pattern. We analyze this algorithm and provide its pseudocode in \cref{ch:coloringSquareFree}.

### 2020-09-04 paper/perfectGraphs.tex
Implementation WIP
Work on combinatorial graphs

\begin{theorem}[Perfect graph theorem]
	A graph is perfect if and only if its complement graph is also perfect. \todo{Should we give some proof of that here? Maybe based on proof in \cite{GC03}}
\end{theorem}

@@

In 1961 Berge conjured that a graph is perfect iff it contains no odd hole and no odd antihole in what has become known as a strong perfect graph conjecture. In 2001 Chudnovsky et al. have proven it and published the proof in an over 150 pages long paper \citetitle{MC06} \cite{MC06}. The following overview of the proof will be based on this paper and on an article with the same name by Cornuéjols \cite{GC03}.

\begin{theorem}[Strong perfect graph theorem]
	A graph is perfect if and only if it is Berge.
\end{theorem}

\TODO{How long and detailed overview of the proof should we provide?}
\TODO{Znakomite pytanie. Generalnie ponieważ dowód jest drobiazgowy i trikowy, to wystarczy "z lotu ptaka" tj. to, co Chudnovsky i Seymour piszą w pracach popularnych i referują w wystąpieniach gościnnych typu te nagrania na YT. Głębiej nie ma sensu. Warto podkreślić, że technika dowodu i sam algorytm są zależne na dużo głębszym poziomie niż widać tj. ktoś mając dowód Stron Perfect Graph Theorem pewnie nie rozgryzłby algorytmu i vice versa.}

@@

The following is based on the paper by Maria Chudnovsky et al. \citetitle*[]{MC05}. We will not provide full proof of its correctness, but will aim to show the intuition behind the algorithm.

Berge graph recognition algorithm could be divided into two parts: first we check if either $G$ or $\overline{G}$ contain any of a number of simple structures as a induced subgraph (\ref{SimpleStructures}). If they do, we output that graph is not Berge and stop. Else, we check if there is a near-cleaner for a shortest odd hole (\ref{AmenableHoles}).

\subsection{Simple structures}
\label{SimpleStructures}

@@

% \begin{floatingfigure}{0.35\textwidth}

\begin{wrapfigure}{r}{0.35\textwidth}
	\input{tikzpictures/pyramid.tex}
	\caption{An example  of a pyramid.}%
	\vspace{-1.2cm}
	% \end{floatingfigure}
\end{wrapfigure}

@@

First, let us enumerate all 6-tuples $b_1, b_2, b_3, s_1, s_2, s_3$ such that:
\begin{itemize}
	\item $\{b_1, b_2, b_3\}$ is a triangle
	\item for $1 \leq i < j \leq 3$, ${b_i, s_i}$ is disjoint from ${b_j, s_j}$ and $b_ib_j$ is the only edge between them
	\item there is a vertex $a$ adjacent to all of $s_1, s_2, s_3$ and to at most one of $b_1, b_2, b_3$, such that if $a$ is adjacent to $b_i$, then $s_i \ b_i$.
\end{itemize}

There are $O(|V(G)|^6)$ 6-tuples, and it takes $O(|V(G)|)$ time to check each one. For each such 6-tuple we follow with the rest of the algorithm.

We define $M = V(G) \setminus \{b_1, b_2, b_3, s_1, s_2, s_3\}$. Now, for each $m \in M$, we set $S_1(m)$ equal to the shortest path between $s_1$ and $m$ such that $s_2, s_3, b_2, b_3$ have no neighbors in its interior, if such a path exists. We set $S_2$ and $S_3$ similarly. Then similarly we set $T_1(m)$ to be the shortest path between $m$ and $b_1$, such that $s_2, s_3, b_2, b_3$ have no neighbors in its interior, if such a path exists. We do similar for $T_2$ and $T_3$. It takes $O(|V(G)|^2)$ time to calculate paths $T_i(m)$ for all $i$ and $m$.

Now, we will calculate all possible paths $P_i$. For each $m \in M \cup \{b_1\}$ we will define a path $P_1(m)$ and paths $P_2(m)$, $P_3(m)$ will be defined in a similar manner.

@@

If $s_1 \neq b_1$, then $P_1(b_1)$ is undefined and for all $m \in M$ we will check if all the following are true:
\begin{itemize}
	\item $m$ is nonadjacent to all of $b_2, b_3, s_2, s_3$
	\item $S_1(m)$ and $T_1(m)$ both exist
	\item $V(S_1(m) \cap T_1(m)) = \{m\}$
	\item there are no edges between $V(S_1(m) \setminus m)$ and $V(T_1(m) \setminus m)$
\end{itemize}
If so, then we assign a path $s_1-S_1(m)-m-T_1(m)-b_1$ to $P_1(m)$, otherwise we let $P_1(m)$ be undefined. It takes $O(|V(G)|^2)$ to check this, given $m$. We assign $P_2$ and $P_3$ in a similar manner. Total time of finding all $P_i(m)$ paths for a given 6-tuple is $O(|V(G)|^3)$.

@@

For $1 \leq i < j \leq 3$, we say that $(m_i, m_j)$ is a \emph{good $(i, j)$-pair}, iff $m_i \in M \cup \{b_i\}$, $m_j \in M \cup \{b_j\}$, $P_i(m_i)$, $P_j(m_j)$ both exist and the sets $V(P_i(m_i))$,$V(P_j(m_j))$ are both disjoint and $b_ib_j$ is the only edge between them.

@@

If after examining all choices of $b_1, b_2, b_3, s_1, s_2, s_3$ we find no pyramid, output that $G$ contains no pyramid. Since there are $O(|V(G)|^6)$ such choices and it takes a time of $O(|V(G)|^3)$ to analyze each one, the total time is $O(|V(G)|^9)$.

\TODO{some proofs}

@@

Most obvious way to find a jewel would be to enumerate all choices of $v_1, \ldots v_5$, check if a choice is correct and if it is try to find a path $P$ as required. This gives us a time of $O(|V|^7)$. We could speed it up to $O(|V|^6)$ with more careful algorithm, but since whole algorithms takes time $O(|V|^9)$ and our testing showed that time it takes to test for jewels is negligible we decided against it.

@@

A configuration of type $\T_1$ is a hole of length 5. To find it we simply iterate all choices of paths of length of 4, and check if there exists a fifth vertex to complete the hole. See \cref{Optimizations} for more implementation details.

@@

To prove that existence of $\T_2$ configuration implies that the graph is not berge, we will need the following Roussel-Rubio lemma:

@@

@@

Then, for each choice of $v_4$ that is adjacent to $v_1$ and not to $v_2$ and $v_5$ and has a neighbor in $F$ (call it $v_6$) and a nonneibhbor in $X$, we test whether there is a vertex $v_3$, adjacent to $v_2, v_4, v_5$ and not to $v_1$, with a nonneibhbor in $X$. If there is such a vertex $v_3$, find $P$ -- a path from $v_6$ to $v_5$ with interior in $F'$ and return that $v_1, \ldots v_6, P, X$ is a configuration of type $\T_3$. If we exhaust our search and find none, report that graph does not contain it.

To see that the algorithm below has a running time of $O(|V(G)|^6)$, let us note that for each triple $v_1, v_2, v_5$ we examine, of which there are $O(|V(G)|^3)$, there are linear many choices of $X$, each taking $O(|V(G)|^2)$ time to process and generating a linear many choices of $v_4$ which take a linear time to process in turn. This gives us the total running time of $O(|V(G)|^6)$.

We will skip the proof that each graph containing $\T_3$ is not Berge. See section6 6.7 of \cite{MC05} for the proof.

@@

\begin{alg}[Test possible near clener, 5.1 of \cite{MC05}]
	\label{alg:testNearCleaner}
	Input: A graph $G$ containing no simple bad structure \todo{define this somewhere}, and a subset $X \subseteq V(G)$.

>>>>

\begin{theorem}[Perfect graph theorem]
	\label{thm:pgt}
	A graph is perfect if and only if its complement graph is also perfect.
\end{theorem}

This theorem is a consequence of a stronger result proven by \Lovasz:
\begin{theorem}
	\label{thm:omegaalpha}
	A graph $G$ is perfect iff for every induced subgraph $H$, the number of vertices of $H$ is at most $\alpha(H)\omega(H)$. \todo{proof?}
\end{theorem}

Then, since $\alpha(H) = \omega(\overline{H})$ and $\omega(H) = \alpha(\overline{H})$ \cref{thm:omegaalpha} implies \cref{thm:pgt}.

@@

In 1961 Berge conjured that a graph is perfect iff it contains no odd hole and no odd antihole in what has become known as a strong perfect graph conjecture. In 2001 Chudnovsky et al. have proven it and published the proof in an over 150 pages long paper \citetitle{MC06} \cite{MC06}.

\begin{theorem}[Strong perfect graph theorem]
	\label{thm:spgt}
	A graph is perfect if and only if it is Berge.
\end{theorem}

The proof is long and complicated. Moreover, it has little noticeable connection to the algorithm of recognizing Berge graphs we discuss later. Therefore we will discuss it very briefly following the overview by Cornuéjols \cite{GC03}.

\subsubsection{Basic classes of perfect graphs}
Bipartite graphs are perfect, since we can color them with two colors. From the theorem of König we get that line graphs of bipartite graphs are also perfect \cite{Knig1916, GC03}. From the perfect graph theorem (\cref{thm:pgt}) it follows that complements of bipartite graphs and complement of line graphs of bipartite graphs are also perfect. We will call these four classes \emph{basic}.

\subsubsection{2-join, Homogeneous Pair and Skew Partition}
A graph $G$ has a \emph{2-join} if its vertices can be partitioned into sets $V_1$, $V_2$, each of size at least three, and there are nonempty disjoint subsets $A_1, B_1 \subseteq V_1$ and $A_2, B_2 \subseteq V_2$, such that all vertices of $A_1$ are adjacent to all vertices of $A_2$, all vertices of $B_1$ are adjacent to all vertices of $B_2$, and these are the only edges between $V_1$ and $V_2$. When a graph $G$ has a 2-join, it can be decomposed onto two smaller graphs $G_1$, $G_2$, so that $G$ is perfect iff $G_1$ and $G_2$ are perfect \cite{Cornujols1985}.

A graph $G$ has a \emph{homogeneous pair} if $V(G)$ can be partitioned into subsets $A_1$, $A_2$, $B$, such that $|A_1|+|A_2| \geq 3$, $|B| \geq 2$ and if a vertex $v \in B$ is adjacent to a vertex from $A_i$, then it is adjacent to all vertices from $A_i$. Chvátal and Sbihi proved that no minimally imperfect graph has a homogeneous pair \cite{Chvtal1987}.

A graph $G$ has a \emph{skew partition} if $V(G)$ can be partitioned into nonempty subsets $A, B, C, D$ such that there are all possible edges between $A$ and $B$ and no edges between $C$ and $D$. Chudnovsky et at. proved that no minimally imperfect graph has a skew partition.

The proof of \cref{thm:spgt} is a consequence of the \emph{Decomposition theorem}:
\begin{theorem}[Decomposition theorem]
	\label{thm:decomposition}
	Every Berge graph $G$ is basic or has a skew partition or a homogeneous pair, or $G$ or $\overline{G}$ has a 2-join.
\end{theorem}

See \cite{MC06} for the proof of \cref{thm:decomposition} and \cref{thm:spgt}.

@@

The following is based on the paper by Maria \citeauthor{MC05} \citetitle*[]{MC05} \cite{MC05}. We will not provide full proof of its correctness, but will aim to show the intuition behind the algorithm.

Berge graph recognition algorithm could be divided into two parts: first we check if either $G$ or $\overline{G}$ contain any of a number of simple forbidden structures (\cref{SimpleStructures}). If they do, we output that graph is not Berge and stop. Else, we check if there is a near-cleaner for a shortest odd hole (\cref{AmenableHoles}).

\subsection{Simple forbidden structures}
\label{SimpleStructures}

@@

\begin{wrapfigure}{r}{0.35\textwidth}
	\input{tikzpictures/pyramid.tex}
	\caption{An example  of a pyramid.}%
	\vspace{-1.2cm}
\end{wrapfigure}

@@

\begin{alg}[Test if $G$ contains a Pyramid]
	\label{alg:testPyramid}
	Input: A graph $G$.

	\noindent Output: Returns whether $G$ contains a pyramid as an induced subgraph.
\end{alg}
\begin{algtext}

Enumerate all 6-tuples $(b_1, b_2, b_3, s_1, s_2, s_3)$ such that:
\begin{itemize}
	\item $\{b_1, b_2, b_3\}$ is a triangle
	\item for $1 \leq i < j \leq 3$, $\{b_i, s_i\}$ is disjoint from $\{b_j, s_j\}$ and $b_ib_j$ is the only edge between them
	\item there is a vertex $a$ adjacent to all of $s_1, s_2, s_3$ and to at most one of $b_1, b_2, b_3$, such that if $a$ is adjacent to $b_i$, then $s_i = b_i$.
\end{itemize}

There are $O(|V(G)|^6)$ 6-tuples, and for each we check above conditions in $O(|V(G)|)$ time. For each such 6-tuple we follow with the rest of the algorithm.

Let $M \leftarrow V(G) \setminus \{b_1, b_2, b_3, s_1, s_2, s_3\}$.

For each $m \in M$, set $S_1(m)$ equal to the shortest path between $s_1$ and $m$ such that $s_2, s_3, b_2, b_3$ have no neighbors in its interior, if such a path exists. Set $S_2$ and $S_3$ similarly. Set $T_1(m)$ to be the shortest path between $m$ and $b_1$, such that $s_2, s_3, b_2, b_3$ have no neighbors in its interior, if such a path exists, and set $T_2$, $T_3$ similarly. It takes $O(|V(G)|^2)$ time to calculate paths $S_i(m)$ and $T_i(m)$ for all $i$ and $m$.

Now, we will calculate all possible paths $P_i$. For each $m \in M \cup \{b_1\}$ we will define a path $P_1(m)$ with paths $P_2(m)$, $P_3(m)$ defined in a similar manner.

@@

If $s_1 \neq b_1$, then $P_1(b_1)$ is undefined and for all $m \in M$ we will check if all the following are true:
\begin{itemize}
	\item $m$ is nonadjacent to all of $b_2, b_3, s_2, s_3$
	\item $S_1(m)$ and $T_1(m)$ both exist
	\item $V(S_1(m) \cap T_1(m)) = \{m\}$
	\item there are no edges between $V(S_1(m) \setminus m)$ and $V(T_1(m) \setminus m)$
\end{itemize}
If it is true for some $m \in M$, then we assign $P_1(m) \leftarrow s_1-S_1(m)-m-T_1(m)-b_1$, otherwise we let $P_1(m)$ be undefined. It takes $O(|V(G)|^2)$ to check this, for a given $m$. We assign $P_2$ and $P_3$ in a similar manner. Total time of finding all $P_i(m)$ paths for a given 6-tuple is $O(|V(G)|^3)$.

@@

For $1 \leq i < j \leq 3$, we say that $(m_i, m_j)$ is a \emph{good $(i, j)$-pair}, iff $m_i \in M \cup \{b_i\}$, $m_j \in M \cup \{b_j\}$, $P_i(m_i)$ and $P_j(m_j)$ both exist, and the sets $V(P_i(m_i))$,$V(P_j(m_j))$ are both disjoint and $b_ib_j$ is the only edge between them.

@@

If after examining all choices of $b_1, b_2, b_3, s_1, s_2, s_3$ we find no pyramid, output that $G$ contains no pyramid. Since there are $O(|V(G)|^6)$ such choices and it takes a time of $O(|V(G)|^3)$ to analyze each one, the total time is $O(|V(G)|^9)$.
\end{algtext}

It is easy to see, that if algorithm outputs that $G$ contains a pyramid, it indeed does. The proof of the converse is rather technical and we refer to \cite{MC05} for it.

@@

Most obvious way to find a jewel would be to enumerate all (possibly chordal) cycles of length 5 as $v_1, \ldots v_5$, check if it has all required nonedges and if it does, try to find a path $P$ as required. This gives us a time of $O(|V|^7)$. We could speed it up to $O(|V|^6)$ with more careful algorithm, but since whole Berge recognition algorithm takes time $O(|V|^9)$ and our testing showed that time it takes to test for jewels is negligible we decided against it.

@@

A configuration of type $\T_1$ is a hole of length 5. To find it, we simply iterate over all paths of length 4 and check if there exists a fifth vertex to complete the hole. See \cref{sec:usesGeneration} for more implementation details.

@@

To prove that existence of $\T_2$ configuration implies that the graph is not Berge, we will need the following Roussel-Rubio lemma:

@@

\begin{alg}
	\label{alg:t3}
	Input: A graph $G$.

	\noindent Output: Returns whether $G$ contains a configuration of type $T_3$ as an induced subgraph.
\end{alg}
\begin{algtext}

@@

Then, for each choice of $v_4$ that is adjacent to $v_1$ and not to $v_2$ and $v_5$ and has a neighbor in $F$ (call it $v_6$) and a nonneibhbor in $X$, we test whether there is a vertex $v_3$, adjacent to $v_2, v_4, v_5$ and not to $v_1$, with a nonneibhbor in $X$. If there is such a vertex $v_3$, find $P$ -- a path from $v_6$ to $v_5$ with interior in $F'$ and return that $v_1, \ldots v_6, P, X$ is a configuration of type $\T_3$. If we exhaust our search and find none, report that graph does not contain it.
\end{algtext}

To see that the \cref{alg:t3} has a running time of $O(|V(G)|^6)$, let us note that for each triple $v_1, v_2, v_5$ we examine, of which there are $O(|V(G)|^3)$, there are linear many choices of $X$, each taking $O(|V(G)|^2)$ time to process and generating a linear many choices of $v_4$ which take a linear time to process in turn. This gives us the total running time of $O(|V(G)|^6)$.

We will skip the proof that each graph containing $\T_3$ is not Berge. See section 6 6.7 of \cite{MC05} for the proof.

@@

\begin{alg}[Test possible near clener, 5.1 of \cite{MC05}]
	\label{alg:testNearCleaner}
	Input: A graph $G$ containing no simple forbidden structure, and a subset $X \subseteq V(G)$.
